---
title: "Assessing Amtrak Trespasser Fatalities in California"
subtitle: "DSAN 6750 / PPOL 6805: GIS for Spatial Data Science"
authors:
  - name: Lindsay Strong
    email: ls1568@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
bibliography: GIS_Manuscript.bib
title-block-banner: "#E69F0095"
title-block-banner-color: "black"
format:
  html:
    df-print: kable
  pdf:
    # https://quarto.org/docs/reference/formats/pdf.html
    link-citations: true
prefer-html: true
---

## Introduction

Several previous studies have found robust relationships between spatial properties of a country's **capital city** and that country's propensity for **conflict** and **misgovernance**.

Perceptions of this linkage also have an effect on "coup-proofing" decisions made by national governments. A recent BBC interview with Equatorial Guinea's President Teodoro Obiang, for example, highlighted this as a factor behind his decision to relocate the capital city:

> It's the remoteness of Oyala that makes it so appealing to President Obiang. In a rare interview he described how rebels had recently plotted a seaborne assault on his palace in the current capital, Malabo. 'We need a secure place for my government and for future governments. That's why we have created Oyala, to guarantee the government of Equatorial Guinea.' [@sackur_equatorial_2012]

This case is far from exceptional, as an even more recent *Washington Post* article points out with respect to Myanmar's decision to move its capital from Yangon to Naypyidaw:

> Analysts have described the decision as motivated by a desire to secure the military’s seat of power from any threat of protests or invasions. [@berger_myanmars_2021]

Most of these studies, however, are based on observations of **conflict events**. In this study, we study the more fundamental variable of a capital's distance from the **population centroid** of the country.

## Literature Review

@campante_capital_2019 analyzes the relationship between the location of a **capital city** and the degree of conflict and misgovernance in a given country. Their two key findings are that:

> Conflict is more likely to emerge (and dislodge incumbents) closer to the capital

and

> Isolated capitals are associated with misgovernance.

This first finding is illustrated in @fig-conflict-dist

![](images/conflict_dist.svg){#fig-conflict-dist}

## Methodology

The **population centroids** we use herein might require some explanation, since the term "centroid" can be ambiguous.

Here, the population centroids are drawn from @hall_population_2019

## Exploratory Data Analysis (EDA)

Here we plot the base GIS objects we're analyzing: the location of each **capital city** (in purple) and each **population centroid** (in yellow).

```{r}
# import libraries and csv
cb_palette <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442","#0072B2", "#D55E00", "#CC79A7"
)
library(sf)
library(readr)
library(mapview)
trespasser_df <- read_csv("Trespasser_Fatalities_20241118.csv")
trespasser_sf <- trespasser_df |> st_as_sf(wkt="the_geom", crs=4326)
library(dplyr)
# getting only Amtrak trespasser strikes
trespasser_sf <- trespasser_sf |> filter(RAILROAD == "ATK") |> filter(StateName == "CA")
mapview(trespasser_sf)
```

```{r}
library(tigris)
library(tidycensus)
library(tidyverse)
options(tigris_use_cache = TRUE)
census_api_key("22de4cafe1259cf144997981fc5e359e41276773", install = TRUE, overwrite = TRUE)


california <- get_acs(
  state = "CA",
  geography = "tract",
  variables = c("B01003_001", "B25043_001"),
  geometry = TRUE,
  year = 2020
)

california_tracts <- california |> select(-moe) |>
  pivot_wider(
    names_from = variable,  
    values_from = estimate, 
    names_glue = "{variable}" 
  ) |> mutate(
    population_density = B01003_001 / B25043_001
  ) |> select(-B01003_001, -B25043_001)

trespasser_sf <- st_transform(trespasser_sf, crs=3587)
california_tracts <- st_transform(california_tracts, crs=3587)
#intersecting_indices <- st_intersects(trespasser_sf, california_tracts)
california_tracts <- st_join(california_tracts, trespasser_sf, join = st_contains) |> group_by(GEOID, NAME, geometry, population_density,) |> summarise(strike_count = sum(!is.na(OBJECTID)), .groups = 'drop')

california_tracts_with_strikes <- california_tracts |> filter(strike_count > 0) |> na.omit()
mapview(california_tracts_with_strikes, zcol="strike_count")
```

```{r}
# library(tidycensus)
# library(dplyr)
# library(sf)
# library(tigris)

# census_api_key("22de4cafe1259cf144997981fc5e359e41276773", install = TRUE, overwrite = TRUE)

# library(tidycensus)
# library(tidyverse)
# options(tigris_use_cache = TRUE)

# california <- get_acs(
#   state = "CA",
#   geography = "tract",
#   variables = c("B01003_001", "B25043_001"),
#   geometry = TRUE,
#   year = 2021
# )

# california <- st_transform(california, crs = 4326)

# intersecting_indices <- st_intersects(california, route_sf)
# intersecting_tracts <- california[!sapply(intersecting_indices, is_empty), ]
# intersecting_tracts <- intersecting_tracts %>% dplyr::select(-moe)

# intersecting_tracts <- intersecting_tracts %>%
#   pivot_wider(
#     names_from = variable,  
#     values_from = estimate, 
#     names_glue = "{variable}" 
#   )

# intersecting_tracts <- intersecting_tracts %>%
#   mutate(
#     population_density = B01003_001 / B25043_001
#   )

# intersecting_trespasser_indices<- st_intersects(trespasser_sf, intersecting_tracts)
# california_strikes <- trespasser_sf[!sapply(intersecting_trespasser_indices, is_empty), ]
# tracts_with_strikes <- st_join(intersecting_tracts, california_strikes, join = st_contains, left = TRUE)
# strikes_per_tract <- tracts_with_strikes %>%
#   group_by(GEOID, NAME, geometry, population_density, B01003_001) %>%
#   summarise(strike_count = sum(!is.na(OBJECTID)), .groups = 'drop')
# mapview(strikes_per_tract, zcol="strike_count", layer.name = "Number of Trespasser Fatalities")
```

We then construct an **area-normalized** measure of capital-centroid distance $\text{dist}^{\textsf{AN}}$, using the formula

$$
\text{dist}^{\textsf{AN}}_i = \text{dist}_i / \sqrt{\text{area}_i}.
$$

A plot of this measure by country looks as follows:

```{r}
#| label: plot-area
#merged_area_sf <- readRDS("merged_area_sf.rds")
#mapview(merged_area_sf, zcol="scaled_dist")
```

## Hypothesis Testing (Regression)

```{r}
library(spatstat)
strike_tract_centroids <- st_centroid(california_tracts_with_strikes)
strike_tract_centroids <- st_transform(strike_tract_centroids)
hull_sf <- st_convex_hull(st_union(strike_tract_centroids))
hull_sf <- st_buffer(hull_sf, dist=1000)
```

```{r}
library(dplyr)
pop_sf <- strike_tract_centroids |> dplyr::select(population_density)
pop_sf <- st_transform(pop_sf, crs=3587)
hull_sf <- st_transform(hull_sf, crs=3587)
point_coords <- st_coordinates(pop_sf)
marks_data <- pop_sf$population_density
pop_ppp <- ppp(x = point_coords[, 1], 
               y = point_coords[, 2], 
               marks = marks_data, 
               window = as.owin(hull_sf))
#pop_ppp <- as.ppp(pop_sf, W=as.owin(hull_sf))
plot(density(pop_ppp), main="Density by Centroids")
plot(density(pop_ppp, weights=pop_ppp$marks), main="Density by Population Density")
pop_int <- density(pop_ppp, weights=pop_ppp$marks)
```

```{r}
num_regions <- 3
region_labels <- c("Low", "Medium", "High")
pop_vals <- pop_int
pop_quant <- quantile(pop_vals, probs=(0:num_regions)/num_regions, na.rm=TRUE)
pop_cut <- cut(pop_vals, breaks=pop_quant, labels=region_labels)
pop_areas <- tess(image=pop_cut)
plot(pop_areas)
```

```{r}
strike_ppp <- as.ppp(st_as_sfc(trespasser_sf), Window(pop_ppp))
plot(density(strike_ppp))
```

```{r}
obs_strike_counts <- quadratcount(strike_ppp, tess=pop_areas) |> as.vector()
names(obs_strike_counts) <- region_labels
obs_strike_counts
```

```{r}
set.seed(6805)
compute_quadrat_counts <- function(sim_ppp) {
  sim_counts <- quadratcount(sim_ppp, tess=pop_areas) |> as.vector()
  names(sim_counts) <- region_labels
  return(sim_counts)
}
gen_sims_ppp <- function(num_sims) {
  strike_sims <- spatstat.random::rpoint(
    n = nrow(trespasser_sf),
    f = pop_int,
    nsim = num_sims
  )
  return(strike_sims)
}
full_sims_list <- gen_sims_ppp(num_sims = 999)
full_sim_area_counts <- lapply(X=full_sims_list, FUN=compute_quadrat_counts)
full_count_df <- as_tibble(full_sim_area_counts) |> t() |> as_tibble()
colnames(full_count_df) <- region_labels

mc_df <- bind_rows(full_count_df, obs_strike_counts)
full_count_df |> ggplot(aes(x=High)) +
  #geom_bar(stat='count') +
  geom_density(fill=cb_palette[2], alpha=0.5) +
  geom_vline(xintercept = obs_strike_counts["High"], linetype="dashed", color=cb_palette[1]) +
  theme_classic()
```

```{r}
q4_more_extreme_df <- mc_df[mc_df$High >= obs_strike_counts["High"],]
q4_prop_more_extreme <- nrow(q4_more_extreme_df) / nrow(mc_df)
q4_prop_more_extreme
```

```{r}
mc_df <- bind_rows(full_count_df, obs_strike_counts)
full_count_df |> ggplot(aes(x=Medium)) +
  #geom_bar(stat='count') +
  geom_density(fill=cb_palette[2], alpha=0.5) +
  geom_vline(xintercept = obs_strike_counts["Medium"], linetype="dashed", color=cb_palette[1]) +
  theme_classic()
```

```{r}
q4_more_extreme_df <- mc_df[mc_df$Medium <= obs_strike_counts["Medium"],]
q4_prop_more_extreme <- nrow(q4_more_extreme_df) / nrow(mc_df)
q4_prop_more_extreme
```

```{r}
mc_df <- bind_rows(full_count_df, obs_strike_counts)
full_count_df |> ggplot(aes(x=Low)) +
  #geom_bar(stat='count') +
  geom_density(fill=cb_palette[2], alpha=0.5) +
  geom_vline(xintercept = obs_strike_counts["Low"], linetype="dashed", color=cb_palette[1]) +
  theme_classic()
```

```{r}
q4_more_extreme_df <- mc_df[mc_df$Medium <= obs_strike_counts["Medium"],]
q4_prop_more_extreme <- nrow(q4_more_extreme_df) / nrow(mc_df)
q4_prop_more_extreme
```


```{r}
#| label: plot-regression
#merged_sub_sf <- readRDS("merged_sub_sf.rds")
#merged_sub_sf |> head()
```

```{r}
#merged_sub_sf |> ggplot(aes(x=scaled_dist, y=total_score, label=NAME_ENGLI)) +
#  geom_point() +
#  geom_smooth(method='lm', formula= y~x) +
#  geom_text(size=4, nudge_y = 0.075) +
 # theme_classic()
```

## Discussion

## Conclusion

Our evidence indicates that the spatial dynamics of **conflict** differ from the spatial dynamics of **misgovernance**. Whereas 